Using Fundamental Measure Theory to Treat the Correlation Function of the Inhomogeneous Hard-Sphere Fluid

### 2012-08-15 papers/contact/paper.tex
edit of paper and new contact data for mark II WB Tested everything on kipu.

In order to motivate the derivation of our functional for the value of
the correlation function at contact, we begin by deriving the
well-known value of $g_\sigma$ for a homogeneous hard-sphere fluid
from the Carnahan-Starling free energy.  The contact value of the
correlation function density is most easily found by using the
contact-value theorem, which states that the pressure on any hard
surface is determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact} \\
  &= k_BT n g_\sigma
\end{align}
Since we are interested in the correlation function at the surface of
the hard spheres, all we need compute is the pressure on that surface,
and we'll have our answer.  The pressure on a hard sphere can be
readily computed from the dependence of the Carnahan-Starling free
energy on hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{24} \sigma^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= 2\frac{dA_{HS}}{d\eta} \frac{d\eta}{d\sigma} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{6 \eta}{\sigma} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi \sigma^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 2\pi \sigma^2} \frac{dA_{HS}}{d\sigma} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3} \label{eq:cs-g}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

\section{Fundamental-Measure Theory}

We use the White Bear version of the Fundamental-Measure Theory~(FMT)
functional published in reference~\cite{roth2002whitebear}.  The FMT
functional describes the excess free energy of a hard-sphere fluid.
This particular FMT reduces to the Carnahan-Starling equation of state
for homogeneous systems.
\begin{equation}
A_\textit{HS}[n] = k_B T \int \left(\Phi_1(\rr) + \Phi_2(\rr) + \Phi_3(\rr)\right) d\rr \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
  n_3 + (1-n_3)^2 \ln(1-n_3)
}{
  36\pi n_3^2\left( 1 - n_3 \right)^2
} ,
\end{align}
using the weighted densities
\begin{align}
  n_3(\rr) &= \int n(\rr') \Theta(\left|\rr - \rr'\right| - \sigma/2) d\rr' \\
  n_2(\rr) &= \int n(\rr') \delta(\left|\rr - \rr'\right| - \sigma/2) d\rr'
\end{align}
\begin{align}
  \mathbf{n}_{V2} &= \int n(\rr') \delta(\left|\rr - \rr'\right| - \sigma/2) \frac{\rr - \rr'}{\left|\rr - \rr'\right|} d\rr' \quad \\
  \mathbf{n}_{V1} = \frac{\mathbf{n}_{V2}}{2\pi \sigma} , \quad
  n_1 &= \frac{n_2}{2\pi \sigma} , \quad
  n_0 = \frac{n_2}{\pi \sigma^2}
\end{align}

\derivation{
  \begin{widetext}
}

>>>>

In order to motivate the derivation of our functional for the value of
the correlation function at contact, we begin by deriving the
well-known value of $g_\sigma$ for a homogeneous hard-sphere fluid
from the Carnahan-Starling free energy.  The contact value of the
correlation function density is most easily found by using the
contact-value theorem, which states that the pressure on any hard
surface is determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact} \\
  &= k_BT n g_\sigma
\end{align}
Since we are interested in the correlation function at the surface of
the hard spheres, all we need compute is the pressure on that surface,
and we'll have our answer.  The pressure on a hard sphere can be
readily computed from the dependence of the Carnahan-Starling free
energy on hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{\pi}{6} \sigma^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere diameter:
\begin{align}
  \frac{dA_{HS}}{d\sigma} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{d\sigma} \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{\sigma} \label{eq:dAhsdR}
\end{align}
This derivative gives us half the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by half the total
area, which means dividing by $N 8\pi \sigma^2$, since the relevant
area is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 8\pi \sigma^2} \frac{dA_{HS}}{d\sigma} \\
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the well-known
correlation function evaluated at contact.
\begin{align}
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3} \label{eq:cs-g}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this derivation to the inhomogeneous
fluid requires that we find the pressure on particular spheres---or
potentially particular parts of particular spheres.
